---
title: "Differences from Temporal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differences from Temporal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(zeitig)
```

zeitig follows the [Temporal](https://tc39.es/proposal-temporal/docs/) specification, but all date
and time logic comes from the Rust crate [jiff](https://docs.rs/jiff), which is modeled on Temporal
without claiming to implement it exactly. Code ported from JavaScript works as expected in most
cases. This article lists the differences you are most likely to notice.

## How closely zeitig follows Temporal

The package's tests run about 6,300 operations through both zeitig and the reference Temporal
implementation and compare the results. The operations cover parsing, arithmetic, differences and
rounding with every unit and rounding mode, formatting options, and time zone transitions, and
include a selection of cases from Temporal's official test suite (test262). 98.8% of the cases
give the same result. The remaining 1.2% are the differences described below, mostly dates beyond
the year 9999. No other case may differ, or the tests fail.

zeitig implements 160 of the 216 methods and properties of Temporal's classes (74%). Most of the
rest belong to `PlainYearMonth`, `PlainMonthDay`, calendars and `toLocaleString()`.

## Only the ISO 8601 calendar

Temporal supports many calendars (Hebrew, Islamic, Japanese, ...); jiff, and therefore zeitig,
supports only the ISO 8601 (proleptic Gregorian) calendar. A string with another calendar is an
error rather than being read as ISO 8601:

```{r, error = TRUE}
plain_date("2024-03-15[u-ca=iso8601]")
plain_date("2024-03-15[u-ca=hebrew]")
```

Some other parts of Temporal are missing too:

* There are no `PlainYearMonth` or `PlainMonthDay` types. Use a `plain_date()` and ignore the
  fields you do not need.
* There is no `toLocaleString()`. Use `temporal_strftime()` for custom formats:

```{r}
temporal_strftime(plain_date(2024, 3, 15), "%A %d %B %Y")
```

## A smaller range of dates

Temporal allows instants about 273,790 years either side of 1970. zeitig only covers the years
-9999 to 9999 (instants stop a day or two short of either end, so that any UTC offset can be
applied), and a duration can be at most 19,998 years:

```{r, error = TRUE}
plain_date(9999, 12, 31)
plain_date(10000, 1, 1)
```

## Comparing durations

Temporal durations have no `==` or `<`. In zeitig, `==` compares the fields of two durations, so one
hour is not `==` to 60 minutes. To compare their lengths, use `duration_compare()`:

```{r}
duration(hours = 1) == duration(minutes = 60)
duration_compare(duration(hours = 1), duration(minutes = 60))
```

`<`, `sort()` and `duration_total()` compare by length. Without a `relative_to` date, they treat
days as 24 hours and weeks as 7 days. (Temporal requires `relativeTo` for weeks.) Durations in
months or years have no fixed length, so they need a `relative_to` date:

```{r, error = TRUE}
sort(c(duration(days = 2), duration(hours = 36)))
duration_total(duration(weeks = 2), "hour")
duration(months = 1) < duration(days = 31)
duration_compare(duration(months = 1), duration(days = 31), relative_to = plain_date(2021, 2, 1))
```

In a time zone with daylight saving time, a day is not always 24 hours. Pass a
`zoned_date_time()` as `relative_to` to account for that:

```{r}
z <- zoned_date_time("2020-03-07T12:00-05:00[America/New_York]")
duration_total(duration(days = 1), "hour", relative_to = z)
```

## Comparing zoned date-times

`==`, `<` and `sort()` on zoned date-times compare only the exact time. To also compare the time zone,
as Temporal's `equals()` does, use `temporal_equals()`:

```{r}
utc <- zoned_date_time("2020-01-01T12:00+00:00[UTC]")
london <- with_time_zone(utc, "Europe/London")
utc == london
temporal_equals(utc, london)
```

Like Temporal, `temporal_equals()` treats `UTC`, `Etc/UTC`, `Etc/GMT` and `GMT` as the same zone,
but it does not resolve other alternative names, such as `Asia/Calcutta` for `Asia/Kolkata`. The
same rule decides whether `temporal_until()` and `temporal_since()` may use days or larger units
between two zoned date-times: they need the same zone, and those four names count as one.

## Nanoseconds since the epoch

Temporal returns `epochNanoseconds` as a `BigInt`. R has no 64-bit integer type, and a double
cannot hold nanoseconds exactly, so `epoch_nanoseconds()` returns a string. `instant_from_epoch()`
accepts the same string:

```{r}
x <- instant("2019-03-30T00:45:00.123456789Z")
epoch_nanoseconds(x)
instant_from_epoch(nanoseconds = epoch_nanoseconds(x)) == x
```

`epoch_seconds()` and `epoch_milliseconds()` return doubles, rounded down to whole units.

## UTC offsets with seconds

Before about 1900, many places used local mean time, with UTC offsets that are not whole minutes.
Temporal prints such offsets rounded to the minute; zeitig prints the seconds, so that the string
reads back as the same value. A UTC offset with fractional seconds, which Temporal accepts, is an
error in zeitig:

```{r, error = TRUE}
format(instant("1900-01-01T12:00Z"), time_zone = "Europe/Dublin")
instant("2020-01-01T12:00+01:00:30.5")
```

## R conveniences

These do not change any results, but they have no equivalent in Temporal:

* Every function is vectorised, and each element of a zoned date-time can have its own time zone.
* `+` and `-` with a duration call `temporal_add()` and `temporal_subtract()`. `x - y` for two
  values of the same type is `temporal_since(x, y)`:

```{r}
plain_date(2021, 1, 31) - plain_date(2020, 12, 1)
```
