---
title: "Differences from Temporal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differences from Temporal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(zeitig)
```

zeitig follows the [Temporal](https://tc39.es/proposal-temporal/docs/) specification, but all date
and time logic comes from the Rust crate [jiff](https://docs.rs/jiff), which is modeled on Temporal
without claiming to implement it exactly. Code ported from JavaScript works as expected in most
cases. This article lists the differences you are most likely to notice.

## Only the ISO 8601 calendar

Temporal supports many calendars (Hebrew, Islamic, Japanese, ...); jiff, and therefore zeitig,
supports only the ISO 8601 (proleptic Gregorian) calendar. Some other parts of Temporal are missing
too:

* There are no `PlainYearMonth` or `PlainMonthDay` types. Use a `plain_date()` and ignore the
  fields you do not need.
* There is no `toLocaleString()`. Use `temporal_strftime()` for custom formats:

```{r}
temporal_strftime(plain_date(2024, 3, 15), "%A %d %B %Y")
```

## A smaller range of dates

Temporal allows instants about 273,790 years either side of 1970. zeitig only covers the years
-9999 to 9999, and a duration can be at most 19,998 years:

```{r, error = TRUE}
plain_date(9999, 12, 31)
plain_date(10000, 1, 1)
```

## Comparing durations

Temporal durations have no `==` or `<`. In zeitig, `==` compares the fields of two durations, so one
hour is not `==` to 60 minutes. To compare their lengths, use `duration_compare()`:

```{r}
duration(hours = 1) == duration(minutes = 60)
duration_compare(duration(hours = 1), duration(minutes = 60))
```

`<`, `sort()` and `duration_total()` compare by length. Without a `relative_to` date, they treat
days as 24 hours and weeks as 7 days. (Temporal requires `relativeTo` for weeks.) Durations in
months or years have no fixed length, so they need a `relative_to` date:

```{r, error = TRUE}
sort(c(duration(days = 2), duration(hours = 36)))
duration_total(duration(weeks = 2), "hour")
duration(months = 1) < duration(days = 31)
duration_compare(duration(months = 1), duration(days = 31), relative_to = plain_date(2021, 2, 1))
```

In a time zone with daylight saving time, a day is not always 24 hours. Pass a
`zoned_date_time()` as `relative_to` to account for that:

```{r}
z <- zoned_date_time("2020-03-07T12:00-05:00[America/New_York]")
duration_total(duration(days = 1), "hour", relative_to = z)
```

## Comparing zoned date-times

`==`, `<` and `sort()` on zoned date-times compare only the exact time. To also compare the time zone,
as Temporal's `equals()` does, use `temporal_equals()`:

```{r}
utc <- zoned_date_time("2020-01-01T12:00+00:00[UTC]")
london <- with_time_zone(utc, "Europe/London")
utc == london
temporal_equals(utc, london)
```

## Nanoseconds since the epoch

Temporal returns `epochNanoseconds` as a `BigInt`. R has no 64-bit integer type, and a double
cannot hold nanoseconds exactly, so `epoch_nanoseconds()` returns a string. `instant_from_epoch()`
accepts the same string:

```{r}
x <- instant("2019-03-30T00:45:00.123456789Z")
epoch_nanoseconds(x)
instant_from_epoch(nanoseconds = epoch_nanoseconds(x)) == x
```

`epoch_seconds()` and `epoch_milliseconds()` return doubles, rounded down to whole units.

## R conveniences

These do not change any results, but they have no equivalent in Temporal:

* Every function is vectorised, and each element of a zoned date-time can have its own time zone.
* `+` and `-` with a duration call `temporal_add()` and `temporal_subtract()`. `x - y` for two
  values of the same type is `temporal_since(x, y)`:

```{r}
plain_date(2021, 1, 31) - plain_date(2020, 12, 1)
```
