---
title: "Time zones"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time zones}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(zietig)
```

## Exact time versus wall-clock time

An `instant()` is a point on the global timeline. A `zoned_date_time()` is the same point plus a
time zone, which gives it a wall-clock reading. Plain types have a wall-clock reading but no
position on the timeline.

```{r}
i <- instant("2020-03-08T16:00Z")
with_time_zone(to_zoned_date_time(i, "UTC"), c("America/New_York", "Asia/Tokyo", "+05:30"))
```

Each element carries its own time zone. Identifiers are IANA names (see `available_time_zones()`)
or fixed offsets such as `"+05:30"`.

## Where the database comes from

On Linux and macOS zietig reads the operating system's time zone database
(`/usr/share/zoneinfo`, or the directory in `TZDIR`), so its rules are as current as the system's.
On Windows, which has no such database, the copy of the IANA database bundled with the jiff crate
is used. Setting `ZIETIG_TZDIR` before the first time zone lookup in a session points zietig at
another zoneinfo directory.

The default time zone, used by `now_zoned_date_time()` and for `POSIXct` values without a `tzone`,
is the `TZ` environment variable when set, otherwise `Sys.timezone()`:

```{r}
now_time_zone()
```

## Daylight saving time

When clocks spring forward, some local times do not exist; when they fall back, some occur twice.
`disambiguation` decides which exact time such a wall-clock time means:

```{r}
# 02:30 did not exist in New York on 2019-03-10
gap <- plain_date_time("2019-03-10T02:30")
for (d in c("compatible", "earlier", "later")) {
  print(to_zoned_date_time(gap, "America/New_York", disambiguation = d))
}
```

```{r, error = TRUE}
to_zoned_date_time(gap, "America/New_York", disambiguation = "reject")
```

Arithmetic with calendar units follows the wall clock, while time units are exact:

```{r}
z <- zoned_date_time("2020-03-07T12:00-05:00[America/New_York]")
z + duration(days = 1)
z + duration(hours = 24)
hours_in_day(z + duration(days = 1))
temporal_until(z, z + duration(days = 1))
```

Changing a field of a zoned date-time inside a DST overlap keeps its UTC offset when possible
(`offset = "prefer"`):

```{r}
o <- zoned_date_time("2019-11-03T01:30-05:00[America/New_York]")
temporal_with(o, minute = 45)
temporal_with(o, minute = 45, offset = "ignore")
```

Transitions can be found directly:

```{r}
time_zone_transition(z, "next")
start_of_day(zoned_date_time("2020-03-29T12:00+03:00[Asia/Beirut]"))
```

## Parsing strings with offsets

A string like `2020-01-01T00:00+01:00[America/New_York]` is inconsistent: New York is not at
`+01:00` on that date. By default that is an error; `offset` chooses another resolution.

```{r, error = TRUE}
s <- "2020-01-01T00:00+01:00[America/New_York]"
zoned_date_time(s)
zoned_date_time(s, offset = "use")
zoned_date_time(s, offset = "ignore")
```
