---
title: "Introduction to zudate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to zudate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

zudate brings the [TC39 Temporal](https://tc39.es/proposal-temporal/docs/) date and time model to
R. Every Temporal type is a vectorised R class, and all date/time logic is delegated to the Rust
crate [jiff](https://docs.rs/jiff), which implements Temporal's semantics.

```{r}
library(zudate)
```

## The types

| Temporal              | zudate                | What it represents                                   |
|-----------------------|-----------------------|------------------------------------------------------|
| `Temporal.PlainDate`  | `plain_date()`        | a calendar date, no time or time zone                |
| `Temporal.PlainTime`  | `plain_time()`        | a wall-clock time, no date or time zone              |
| `Temporal.PlainDateTime` | `plain_date_time()` | a date and a wall-clock time, no time zone          |
| `Temporal.Instant`    | `instant()`           | an exact point in time (nanosecond precision)        |
| `Temporal.ZonedDateTime` | `zoned_date_time()` | an exact time plus the time zone used to view it    |
| `Temporal.Duration`   | `duration()`          | a length of time in calendar and clock units         |
| `Temporal.Now`        | `now_instant()`, ...  | the current time                                     |

Each constructor accepts components or [RFC 9557](https://www.rfc-editor.org/rfc/rfc9557)
strings:

```{r}
plain_date(2006, 8, 24)
plain_time(c("03:24:30", "15:23:30.123456789"))
plain_date_time("1995-12-07T03:24:30")
instant("1969-07-20T20:17Z")
zoned_date_time("1995-12-07T03:24:30-08:00[America/Los_Angeles]")
duration(hours = 1, minutes = 30)
```

They are ordinary vectors: they recycle, hold `NA`, live in data frames and survive `saveRDS()`.

```{r}
d <- plain_date(2024, 1:4, 31)
d
data.frame(date = d, weekday = day_of_week(d))
sort(c(d, NA), na.last = TRUE)
```

Out-of-range components are constrained by default, as in Temporal; use `overflow = "reject"` to
get an error instead.

```{r, error = TRUE}
plain_date(2021, 2, 31)
plain_date(2021, 2, 31, overflow = "reject")
```

## Fields

```{r}
x <- plain_date_time("2021-01-03T15:23:30.123456789")
year(x)
week_of_year(x)
year_of_week(x)
c(millisecond(x), microsecond(x), nanosecond(x))
temporal_fields(x)
```

## Arithmetic

Durations are added with `+`/`-` or `temporal_add()`. Adding months clamps the day of month:

```{r}
plain_date(2021, 1, 31) + duration(months = 1)
plain_time(23, 30) + duration(hours = 1)
```

`temporal_until()` and `temporal_since()` compute differences, balanced up to `largest_unit` and
rounded to `smallest_unit`:

```{r}
a <- plain_date(2006, 8, 24)
b <- plain_date(2019, 1, 31)
temporal_until(a, b)
temporal_until(a, b, largest_unit = "year")
b - a
```

Rounding uses Temporal's rounding modes:

```{r}
t <- plain_time(19, 39, 9, 68, 346, 205)
temporal_round(t, "hour")
temporal_round(t, "minute", rounding_increment = 15, rounding_mode = "floor")
temporal_round(duration(minutes = 130), largest_unit = "hour")
duration_total(duration(months = 1), "day", relative_to = plain_date(2020, 2, 1))
```

## Formatting and parsing

`format()` follows Temporal's `toString()` options, and `temporal_strftime()` /
`temporal_strptime()` use strftime-style patterns:

```{r}
z <- zoned_date_time("2020-01-01T15:23:30.123456789+01:00[Europe/Paris]")
format(z, smallest_unit = "minute")
format(z, fractional_second_digits = 3, time_zone_name = "never")
temporal_strftime(z, "%A %d %B %Y, %H:%M %Z")
temporal_strptime("15/07/2024", "%d/%m/%Y", "plain_date")
```

## Base R interoperability

```{r}
as_plain_date(Sys.Date())
as.Date(plain_date(2020, 2, 29))
as_zoned_date_time(as.POSIXct("2020-07-01 12:00:00", tz = "Europe/Paris"))
as.POSIXct(instant("2020-01-01T00:00:00.25Z"))
as_duration(as.difftime(90, units = "mins"))
```

`POSIXct` holds seconds in a double, so conversions to and from it are rounded to the
microsecond. Use the Temporal types (or `epoch_nanoseconds()`) when you need nanoseconds.

## Differences from Temporal

zudate supports only the ISO 8601 calendar, and a few behaviours follow jiff or R conventions;
they are listed in the package's design notes. The most visible ones:

* `x - y` for two values of the same type is `temporal_since(x, y)`.
* `==` and `<` on zoned date-times compare the exact time only; `temporal_equals()` also compares
  the time zone.
* `==` on durations compares fields (`PT1H` is not `==` to `PT60M`); use `duration_compare()`.
