---
title: "Cookbook: everyday tasks with Temporal types"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This article collects short recipes for the date and time tasks that come up most often, in the
spirit of the [Temporal cookbook](https://tc39.es/proposal-temporal/docs/cookbook.html). Each
recipe uses the type that matches the question being asked; the table below is the quick guide.

| If you have…                                    | use                  |
|-------------------------------------------------|----------------------|
| a calendar date (birthday, due date, holiday)   | `plain_date()`       |
| a wall-clock time (opening hours, alarm)        | `plain_time()`       |
| a date and time with no place (a form input)    | `plain_date_time()`  |
| an exact moment (log entry, event timestamp)    | `instant()`          |
| an exact moment *and* where it happened         | `zoned_date_time()`  |
| a length of time ("3 months", "90 minutes")     | `duration()`         |

```{r}
library(zeitig)
```

All functions are vectorised, so every recipe works the same on one value or on a whole column.

## Getting the current date and time

```{r}
now_instant()                          # exact time, no time zone
now_zoned_date_time("Europe/Berlin")   # exact time seen in a time zone
now_plain_date("Asia/Tokyo")           # today's date in Tokyo
now_plain_time()                       # wall-clock time in the session time zone
now_time_zone()
```

## Parsing strings

Every constructor parses ISO 8601 / RFC 9557 strings. Invalid input is an error that says which
element failed; `NA` stays `NA`.

```{r, error = TRUE}
plain_date(c("2024-02-29", "2024-12-25", NA))
plain_time("08:30")
instant("2024-03-10T07:15:00Z")
zoned_date_time("2024-03-10T08:15:00+01:00[Europe/Berlin]")
duration("P1DT12H")
plain_date(c("2024-02-29", "2023-02-29"))
```

For other layouts, use strftime-style patterns:

```{r}
temporal_strptime(c("25/12/2024", "01/01/2025"), "%d/%m/%Y", "plain_date")
temporal_strptime("Mar 10 2024 3:45 PM", "%b %d %Y %I:%M %p", "plain_date_time")
```

## Building values from parts

```{r}
plain_date(2024, 1:12, 1)                  # first day of every month
plain_time(9, c(0, 30))                    # 09:00 and 09:30
plain_date_time(2024, 6, 1, hour = 18)
zoned_date_time(2024, 6, 1, 18, time_zone = "America/New_York")
duration(hours = 1, minutes = 30)
```

## Taking dates and times apart

```{r}
d <- plain_date_time("2024-07-04T16:45:30")
year(d)
month(d)
day_of_week(d)   # 1 = Monday ... 7 = Sunday
week_of_year(d)
hour(d)
temporal_fields(d)
```

## Date arithmetic

Adding months or years keeps the day of the month when it can and clamps to the end of the month
when it cannot:

```{r}
d <- plain_date(2024, 1, 31)
d + duration(months = 1:3)
d + duration(years = 1, days = 10)
d - duration(weeks = 2)
```

If clamping should be an error instead, ask for it:

```{r, error = TRUE}
temporal_add(d, duration(months = 1), overflow = "reject")
```

## Days between two dates

Subtracting two dates gives a duration in days; `temporal_until()` balances into larger units:

```{r}
start <- plain_date(2024, 1, 15)
end <- plain_date(2024, 12, 25)
end - start
temporal_until(start, end, largest_unit = "month")
duration_total(end - start, "day")
duration_total(end - start, "week")
```

## How old is someone?

```{r}
birthdays <- plain_date(c("1990-02-28", "2000-02-29", "2010-12-31"))
today <- plain_date(2024, 2, 28)
age <- temporal_until(birthdays, today, largest_unit = "year")
age
temporal_fields(age)$years
```

## First and last day of the month

```{r}
d <- plain_date(c("2024-02-10", "2023-02-10", "2024-11-30"))
temporal_with(d, day = 1)
temporal_with(d, day = days_in_month(d))
```

## The next Monday, or the next given weekday

The next occurrence strictly after each date (a Monday maps to the following Monday):

```{r}
next_weekday <- function(date, weekday) {
  ahead <- (weekday - day_of_week(date) + 7) %% 7
  ahead[ahead == 0] <- 7
  date + duration(days = ahead)
}
d <- plain_date(2024, 7, 1:7)
data.frame(date = d, weekday = day_of_week(d), next_monday = next_weekday(d, 1))
```

## Business hours and times of day

Plain times compare and sort as wall-clock times, and arithmetic on them wraps at midnight:

```{r}
opening <- plain_time(9)
closing <- plain_time(17, 30)
t <- plain_time(c("08:59", "12:00", "17:30", "21:15"))
t >= opening & t < closing
plain_time(23, 30) + duration(hours = 2)
temporal_until(opening, closing)
```

## Rounding

```{r}
t <- plain_time("14:37:52.123")
temporal_round(t, "minute")
temporal_round(t, "minute", rounding_increment = 15)
temporal_round(t, "hour", rounding_mode = "floor")

dt <- plain_date_time("2024-07-04T16:45:30")
temporal_round(dt, "day")
```

Durations round too, balancing into larger units:

```{r}
temporal_round(duration(minutes = 135), largest_unit = "hour")
temporal_round(duration(seconds = 5000), "minute", largest_unit = "hour")
```

## Comparing and sorting

Comparison operators, `sort()`, `order()`, `min()`/`max()` and `unique()` follow Temporal's
ordering:

```{r}
x <- plain_date(c("2024-05-01", "2023-12-31", "2024-01-15"))
sort(x)
x > plain_date(2024, 1, 1)
temporal_compare(x, "2024-01-15")
range(x)
```

Zoned date-times compare by the exact time, so the same moment in two zones is equal:

```{r}
berlin <- zoned_date_time("2024-06-01T12:00+02:00[Europe/Berlin]")
tokyo <- with_time_zone(berlin, "Asia/Tokyo")
berlin == tokyo
temporal_equals(berlin, tokyo)   # also compares the time zone
```

## Converting between time zones

```{r}
meeting <- zoned_date_time(2024, 3, 15, 9, time_zone = "America/New_York")
with_time_zone(meeting, c("Europe/London", "Asia/Kolkata", "Australia/Sydney"))
```

A wall-clock time in one place, converted to the exact instant and back to a local reading
elsewhere:

```{r}
local <- plain_date_time("2024-11-03T01:30")
to_zoned_date_time(local, "America/Chicago", disambiguation = "later")
to_instant(to_zoned_date_time(local, "America/Chicago"))
```

## Daily schedules across daylight saving time

Adding days to a zoned date-time keeps the local time, even when the day is 23 or 25 hours long.
Adding hours is exact elapsed time:

```{r}
first <- zoned_date_time("2024-03-08T09:00-05:00[America/New_York]")
first + duration(days = 0:3)
first + duration(hours = 24 * (0:3))
hours_in_day(first + duration(days = 0:3))
```

## Timestamps and epoch values

```{r}
x <- instant_from_epoch(milliseconds = 1720000000000)
x
format(epoch_seconds(x), scientific = FALSE)
format(epoch_milliseconds(x), scientific = FALSE)
epoch_nanoseconds(instant("2024-01-01T00:00:00.123456789Z"))
```

Differences between instants are exact and default to seconds:

```{r}
a <- instant("2024-07-01T10:00:00Z")
b <- instant("2024-07-02T12:30:15.5Z")
b - a
temporal_until(a, b, largest_unit = "hour")
duration_total(b - a, "hour")
```

## Durations

```{r}
d <- duration(c("PT90M", "P1DT2H", "PT45S"))
d
duration_total(d, "minute")
sort(d)
sum_duration <- Reduce(`+`, as.list(d))
sum_duration
abs(-duration(hours = 3))
```

Durations with months or years have no fixed length; give a starting date to measure them:

```{r, error = TRUE}
duration_total(duration(months = 1), "day")
duration_total(duration(months = 1), "day", relative_to = c("2024-01-01", "2024-02-01"))
```

## Formatting for display

```{r}
z <- zoned_date_time("2024-07-04T16:45:30.123456789-04:00[America/New_York]")
format(z)
format(z, smallest_unit = "minute", time_zone_name = "never")
format(z, fractional_second_digits = 3)
temporal_strftime(z, "%A %e %B %Y, %H:%M %Z")
temporal_strftime(plain_date(2024, 7, 4), "%d.%m.%Y")
```

## Working with base R and data frames

Temporal vectors are ordinary vectors, so they work as data frame columns:

```{r}
events <- data.frame(
  id = 1:4,
  when = zoned_date_time(
    c(
      "2024-03-09T23:30-05:00[America/New_York]",
      "2024-03-10T08:00-04:00[America/New_York]",
      "2024-03-10T09:15+01:00[Europe/Paris]",
      "2024-03-11T07:45+09:00[Asia/Tokyo]"
    )
  )
)
events$date_utc <- to_plain_date(with_time_zone(events$when, "UTC"))
events$local_hour <- hour(events$when)
events[order(events$when), ]
summary(events$when)
```

Conversions to and from base classes:

```{r}
as_plain_date(as.Date("2024-02-29"))
as.Date(plain_date(2024, 2, 29))
as_zoned_date_time(as.POSIXct("2024-07-04 12:00:00", tz = "Europe/Paris"))
as.POSIXct(instant("2024-07-04T10:00:00Z"), tz = "UTC")
as_duration(as.difftime(90, units = "mins"))
as_difftime(duration(hours = 36), units = "days")
```

`POSIXct` stores seconds in a double, so conversions through it are rounded to the microsecond;
keep values in zeitig types when nanoseconds matter.
